Sketch Recognition using Particle Swarm Optimization Algorithm

### 2009-04-27 trunk/draft/SketchMaster.tex
sss
change to report

% During the editing process you will not want to view and print the
% whole document. Uncomment the command below so that only the
% chapters that you want to work on are visible.
%\includeonly{ChapterIntroduction}
%\documentstyle[11pt,unhthesis]{report}
\documentclass[11pt,doublespace]{Sketchthesis}

@@

\begin{document}
                                                        % TITLE PAGE
                                                        %======================
\title{Sketch Recognition using Particle Swarm Optimization Algorithm}             % My theses title
\author{Maha Mohammed Nabeel El Meseery}                               % Y name
\prevdegrees{B.S.,Computer  Engineering}       		% Your old degree
\major{Computer Engineering}                            % My current major
\degree{Master of Science}                              % My new degree
\degreemonth{}                                      % When awarded. to be added when known
\degreeyear{}                                      %
%th today or first day of writing  December 26 2005 
%\thesisdate[]                                     % Date of document.
\DOCUMENTtype{Thesis}                                   % or DISSERTATION
\Documenttype{Thesis}                                   % or Dissertation
\documenttype{Thesis}                                   % or dissertation
\maketitle

\providecommand{\tabularnewline}{\\}
                                                        % COPYRIGHT PAGE
                                                        %======================
%\copyrightyear{2009}                                    % Delete these
%\makecopyright                                          % if no copyright
                                                    % APPROVAL                                                         %======================
\thispagestyle{empty}               % Don't do headers or footers.
\newpage                                                             
\begin{center}     
\LARGE  
\textbf{Sketch Recognition using Particle Swarm Optimization Algorithm}\\
\normalsize by      \\ 
\Large  \textbf{ Maha mohamed Nabeel El Meseery}\\
\vspace*{0.4in}        
\normalsize
A Thesis Submitted To the\\
Faculty of Engineering, Cairo University\\
in Partial Fulfillment of the\\
Requirements for the Degree of\\
MASTER OF SCIENCE
in\\
COMPUTER ENGINEERING\\

\end{center}

    Approved by\\                                           
Examining Committee:\\
 \makebox[6in]{\hrulefill}\\ 
Professor Nevin Darwish, Thesis Main Advisor\\
 \makebox[6in]{\hrulefill}\\ 
Professor Samia Mashali,Thesis Advisor \\
 \makebox[6in]{\hrulefill}\\  
Professor Magda Fayek,Thesis Advisor\\ 
 \makebox[6in]{\hrulefill}\\ 
Professor  Ashraf Abd El Wahab, Member\\
 \makebox[6in]{\hrulefill}\\ 
Professor Mohamed Zaki Abd ElMegeed, Member\\
 \makebox[6in]{\hrulefill}\\ 
  % \\   you need...\\
%\makeapproval                                           %

@@

                                                        % ACK. PAGE
                                                        %======================
\begin{acknowledgments}                                 % Delete these if
All praise is due to Allah Who guided me to this. I could not truly have been led aright if
Allah had not guided me.

I would like to convey my sincere appreciation and graditude to my supervisors; Dr. Nevin Darwish, Dr. Magdah Fayek and Dr. Samia Mashali. I would also be very gratuful for dr. Mahmoud fakhr el din for his assitance with my thesis. Thanks for all the time devoted in making constructive comments and suggestions for improving this thesis. 
I thank my family and friends for thier moral and support. I specifically thank, dina said, noha mansour and marwah shafee. Thanks, for being always there when I needed you the most. I hope to express my gratiude to all of you for supproting me and praying for me through all these years. 
 %Finally, I am very grateful to my dear parents, my family, and my friends whom I consider
%as my sisters. I would like to thank particularly Radwa Aboudina, Fatma Nada, Maha
%Nabil, Haidi Badr, and Marwa Kamal. Thank you all for being always there when I needed
%you most. Thank you for believing in me and supporting me through all these years. I think
%without your support and your prayers, none of this work would be accomplished.
%ii
 \end{acknowledgments}                                   %

@@

%------------------------------------------------------------------------------
%  Document body - Place text into individual include files, such as
%                  as "chap1.tex", or replace each "\include{}" statement
%                  with your actual document text.
%------------------------------------------------------------------------------
\begin{abstractpage}          
% ABSTRACT PAGE
%==============================
% Creates the abstract page.
%   Your text goes here. Just
%   follow the rules given in
%   the LaTeX manual on how to
%   enter/format text using
%   LaTeX commands.

@@

Sketch recognition is defined as the process of identifying the symbols the user draws using a single or multiple strokes. Users draw strokes using a pen and the system immediately interprets their strokes into objects that can be easily manipulated. This research uses Particle Swarm Algorithm (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. Two algorithm are used to divide strokes, the first algorithm \textsl{ALgS1} divides stroke to polygons. The second one \textit{AlgS2}, divides stroke to segments of either lines or curves. The research focuses on the effect of PSO segmentations to the sketch recognition systems. The final recognition is achieved using a set of geometrical and global shape properties features. A SVM (support vector machine) classifier is used to correctly identify shapes.

@@

\end{abstractpage}                              % This ends the page.

>>>>

% During the editing process you will not want to view and print the
% whole document. Uncomment the command below so that only the
% chapters that you want to work on are visible.
%\includeonly{ChapterIntroduction}
%\documentstyle[11pt,unhthesis]{report}
\documentclass[11pt,doublespace]{report}%{Sketchthesis}

@@

\begin{document}
                                                        % TITLE PAGE
                                                        %======================
          % My theses title  using Particle Swarm Optimization Algorithm
%\author{Maha Mohammed Nabeel El Meseery}                               % Y name
%\prevdegrees{B.S.,Computer  Engineering}       		% Your old degree
%\major{Computer Engineering}                            % My current major
%\degree{Master of Science}                              % My new degree
%\degreemonth{}                                      % When awarded. to be added when known
%\degreeyear{}                                      %
%%th today or first day of writing  December 26 2005 
%%\thesisdate[]                                     % Date of document.
%\DOCUMENTtype{Thesis}                                   % or DISSERTATION
%\Documenttype{Thesis}                                   % or Dissertation
%\documenttype{Thesis}

   \vspace*{0.375in}
               {\Large\bf\title{Sketch Recognition and Understanding}  }
                                 \vspace*{0.375in}

                                \vspace*{\fill}\par
                                      BY
                                \vspace*{\fill}\par
                               {\Large \author{Maha Mohammed Nabeel El Meseery }}
                                     \par
                               B.Sc.,Computer  Engineering
                                \vspace*{\fill}\par
                                    Thesis
                                \vspace*{\fill}\par
                           Submitted to the Faculty of Engineering Cairo University\\
                         in partial fulfillment of \\
                   the requirements for the degree\ of
                                \vspace*{\fill}\par
                                   Master of Science
                                     \par
                                      in
                                     \par
                                   Master of Science
                                \vspace*{\fill}\par

                                                        % or dissertation
\maketitle

\providecommand{\tabularnewline}{\\}
                                                        % COPYRIGHT PAGE
                                                        %======================
%\copyrightyear{2009}                                    % Delete these
%\makecopyright                                          % if no copyright
                                                    % APPROVAL                                                         %======================
%\thispagestyle{empty}               % Don't do headers or footers.
%\newpage                                                             
%\begin{center}     
%\LARGE  
%\textbf{Sketch Recognition using Particle Swarm Optimization Algorithm}\\
%\normalsize by      \\ 
%\Large  \textbf{ Maha mohamed Nabeel El Meseery}\\
%\vspace*{0.4in}        
%\normalsize
%A Thesis Submitted To the\\
%Faculty of Engineering, Cairo University\\
%in Partial Fulfillment of the\\
%Requirements for the Degree of\\
%MASTER OF SCIENCE
%in\\
%COMPUTER ENGINEERING\\
%
%\end{center}
%          
%    Approved by\\                                           
%Examining Committee:\\
% \makebox[6in]{\hrulefill}\\ 
%Professor Nevin Darwish, Thesis Main Advisor\\
% \makebox[6in]{\hrulefill}\\ 
%Professor Samia Mashali,Thesis Advisor \\
% \makebox[6in]{\hrulefill}\\  
%Assoc. Professor Magda Fayek,Thesis Advisor\\ 
% \makebox[6in]{\hrulefill}\\ 
%Professor  Ashraf Abd El Wahab, Member\\
% \makebox[6in]{\hrulefill}\\ 
%Professor Mohamed Zaki Abd ElMegeed, Member\\
% \makebox[6in]{\hrulefill}\\ 
  % \\   you need...\\
%\makeapproval                                           %

@@

                                                        % ACK. PAGE
                                                        %======================
 \def\acknowledgments
   {\vspace*{0.5in}
    \thispagestyle{plain}
    \chapter*{Acknowledgments}
    \addcontentsline{toc}{section}{Acknowledgments}

   }

\def\endacknowledgments{\par}

\begin{acknowledgments}     
                            % Delete these if
All praise is due to Allah Who guided me to this. I could not truly have been led aright if Allah had not guided me.

I would like to convey my sincere appreciation and gratitude to my supervisors; Dr. Nevin Darwish, Dr. Magdah Fayek and Dr. Samia Mashali. I would also be very grateful for Dr. Mahmoud Fakhr El Din for his assistance with my thesis. Thanks for all the time devoted in making constructive comments and suggestions for improving this thesis.

 I thank my family and friends for their moral and support. I specifically thank, dina said, noha mansour and marwah shafee. Thanks, for being always there when I needed you the most. I hope to express my gratitude to all of you for supporting me and praying for me through all these years.
 %Finally, I am very grateful to my dear parents, my family, and my friends whom I consider
%as my sisters. I would like to thank particularly Radwa Aboudina, Fatma Nada, Maha
%Nabil, Haidi Badr, and Marwa Kamal. Thank you all for being always there when I needed
%you most. Thank you for believing in me and supporting me through all these years. I think
%without your support and your prayers, none of this work would be accomplished.
%ii
 \end{acknowledgments}                                   %

@@

%------------------------------------------------------------------------------
%  Document body - Place text into individual include files, such as
%                  as "chap1.tex", or replace each "\include{}" statement
%                  with your actual document text.
%------------------------------------------------------------------------------
\begin{abstract}          
% ABSTRACT PAGE
%==============================
% Creates the abstract page.
%   Your text goes here. Just
%   follow the rules given in
%   the LaTeX manual on how to
%   enter/format text using
%   LaTeX commands.

@@

Sketch recognition is defined as the process of identifying the symbols the user draws using a single or multiple strokes. Users draw strokes using a pen and the system immediately interprets their strokes into objects that can be easily manipulated. This research uses Particle Swarm Algorithm (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. Two algorithm are used to divide strokes, the first algorithm \textit{ALgS1} divides stroke to polygons. The second one \textit{AlgS2}, divides stroke to segments of either lines or curves. The research focuses on the effect of PSO segmentations to the sketch recognition systems. The final recognition is achieved using a set of geometrical and global shape properties features. A SVM (support vector machine) classifier is used to correctly identify shapes.

@@

\end{abstract}                              % This ends the page.

### 2009-04-27 trunk/draft/chapterSurvey.tex
change to report

%In this section, we review three different types of symbol recognition methods.
 Symbol recognition can be based on: strokes, global shape properties, and appearance \cite{Oltmans07}. The role of which each stroke plays in the structure of the sketched symbol is used to determine shapes in stroke based methods. In global shape properties approach, systems investigate general shape properties and its underlying strokes. Using shape properties relaxes the assumption and role of each stroke in the symbol but it does not truly represent the appearance of the symbol. Unlike appearance based methods where individual strokes are disregarded and they focus on the overall appearance of shapes. We choose to use the method based on shape property but added some geometrical and structure properties used in most stroke based approaches. Our method tries to avoid many of the problems faced by approaches based on individual strokes or global shape properties while trying to gain the additional information that the stroke based system use. In the next sections, we look into three different symbol recognition approaches, describe their advantages and challenges.

@@

One of the first systems in gestures recognition is Rubine's algorithm \cite{gestureexample12} . He used Gaussian statistical classifier to classify gestures.  Simple features as length of stroke, smoothness of stroke, the angle that the first part of the gesture is drawn at and properties of the bounding box of the stroke are used to train the classifier to recognize symbols. In our system, we combine Rubine features with others to include the information on how the user draws sketch while recognizing the symbols. In \cite{Long00}, Long shows how similarities between gestures can be analyzed.  He used this information to identify gestures that are likely to confuse the recognition system. This helped interface designers to choose gestures that are easily recognized.

The main problem with gesture based recognition systems that they impose several constrains on how the user can draw. Gestures must be drawn using a specific order and direction to be recognized correctly. Moreover, in most cases gestures do not necessarily look like the symbols they represent (e.g. to represent a rectangle the user draws only the left edge of it). Because of these reasons, gestures recognition systems are not suitable for free hand drawn sketches.

@@

 Systems that employ such representation must first pre-process strokes to break them into geometric primitives (see section \ref{sec:preprocessing} for a review on preprocessing methods). Recognition can then be treated as sub graph matching problem.

@@

 Several visual grammar were introduced \cite{statisticalparsing26,Ladder30,GenericHMM28}. \cite{Ladder30,GenericHMM28}  Used visual description language which will generate the information of the structure and constrain of each object classified. The system automatically generates Java code for the recognizer for each object. The generated code cycles through the strokes drawn to check if any of the recognizers generated had interpreted any symbols partially or completely.  These systems are less computational expensive than graph matching approach but they also operate based on the output of the segmentation process. Any error in the segmentation process leads to huge error while parsing the whole sketch. Moreover, symbols cannot share strokes between them which influence and restrict how the user draws symbols. 
 % complexity of the graph matching problem 
\begin{figure}
	\centering
		\subfigure[An Arrow]
			{\label{fig:arrow1}\includegraphics[scale=0.75]{../../neededfiles/Figures/arrowladderdef1.eps}}
			\hfill
		\subfigure[Arrow grammar definition]%{\cite{Ladder30}}
		 {\label{fig:arr2}\includegraphics[scale=0.88]{../../neededfiles/Figures/arrowladderdef2.eps}}
	\caption[Arrow Definition]{\cite{Ladder30}}
		\label{fig:arrowladderdef1}
\end{figure}

@@

Image processing is one of the disciplines that used Swarm Intelligence. Ant colony was widely used to segment objects in \cite{antSegmentation,antthreshold,SwarmMedical}. Since 1995 when Kennedy and Eberhart \cite{PSOFirst,PSO2} introduced particle swarm algorithm (PSO) the algorithm had been used in various image processing system as an optimization  algorithm \cite{SwarmImageProcessing,Swarmimageprocessing2,Colorquantization,SwarmMedical}.  PSO algorithm was used in \cite{ToADD} to classify pixel to simplify the object segmentation process.

>>>>

%In this section, we review three different types of symbol recognition methods.
 Oltmans\cite{Oltmans07} mentions that symbol recognition can be based on: strokes, global shape properties, and appearance or image based\cite{Oltmans07}. The role of which each stroke plays in the structure of the sketched symbol is used to determine shapes in stroke based methods.  Image or appearance based methods focuses focus on the overall appearance of shapes rather than individual strokes. Unlike global shape properties approach, systems investigate general shape properties. Using shape properties loosens the assumption and role of each stroke in the symbol but it does not truly represent the symbol visually.

 We choose to use the method based on shape property but added some geometrical and structure properties used in most stroke based approaches. Our method tries to avoid many of the problems faced by approaches based on individual strokes or global shape properties while trying to gain the additional information that the stroke based system use. In the next sections, we look into three different symbol recognition approaches, describe their advantages and challenges.

@@

One of the first systems in gestures recognition is Rubine's algorithm \cite{gestureexample12} . He used Gaussian statistical classifier to classify gestures.  Simple features as length of stroke, smoothness of stroke, the angle that the first part of the gesture is drawn at and properties of the bounding box of the stroke are used to train the classifier to recognize symbols. In our system, we combine Rubine features with others to include the information on how the user draws sketch while recognizing the symbols. %In \cite{Long00}, Long shows how similarities between gestures can be analyzed.  He used this information to identify gestures that are likely to confuse the recognition system. This helped interface designers to choose gestures that are easily recognized.

The main problem with gesture based recognition systems that they impose several constrains on how the user can draw. Gestures must be drawn using a specific order and direction to be recognized correctly. Moreover, in most cases gestures do not necessarily look like the symbols they represent (e.g. to represent a rectangle the user draws only the left edge of it). Because of these restrictions, gestures recognition systems are not suitable for free hand drawn sketches.

@@

 Systems that employ such representation must first pre-process strokes to break them into geometric primitives ( section \ref{sec:preprocessing} reviews on preprocessing methods). Recognition can then be treated as sub graph matching problem or constrain satisfaction problem.

@@

 Several visual grammar were introduced \cite{statisticalparsing26,Ladder30,GenericHMM28}. \cite{Ladder30,GenericHMM28}  Used visual description language which will generate the information of the structure and constrain of each object classified. The system automatically generates Java code for the recognizer for each object. The generated code cycles through the strokes drawn to check if any of the recognizers generated had interpreted any symbols partially or completely.  These systems are less computational expensive than graph matching approach but they also operate based on the output of the segmentation process. Any error in the segmentation process leads to huge error while parsing the whole sketch. Moreover, symbols cannot share strokes between them which influence and restrict how the user draws symbols. 
 % complexity of the graph matching problem 
\begin{figure}
	\centering
		\subfigure[An Arrow]
			{\label{fig:arrow1}\includegraphics[scale=0.75]{../../neededfiles/Figures/arrowladderdef1.eps}}
			\hfill
		\subfigure[Arrow grammar definition]%{\cite{Ladder30}}
		 {\label{fig:arr2}\includegraphics[scale=0.88]{../../neededfiles/Figures/arrowladderdef2.eps}}
	\caption[Arrow Definition]{Arrow Definition in \cite{Ladder30}}
		\label{fig:arrowladderdef1}
\end{figure}

@@

Image processing is one of the disciplines that used Swarm Intelligence. Ant colony was widely used to segment objects in \cite{antSegmentation,antthreshold,SwarmMedical}. Since 1995 when Kennedy and Eberhart \cite{PSOFirst,PSO2} introduced particle swarm algorithm (PSO) the algorithm had been used in various image processing system as an optimization  algorithm \cite{SwarmImageProcessing,Swarmimageprocessing2,Colorquantization,SwarmMedical}.  PSO algorithm was used in \cite{PSOIMageClassification} to classify pixel to simplify the object segmentation process.
